---
title: "Reproducible Research Peer Assessment 1"
author: "Alan C Bonnici"
date: "Friday, August 08, 2014"
output: html_document
---

Activity Monitoring Data Analysis
=================================

What follows is a data analysis of an activity monitoring device. The data are data collected by an activity monitoring device worn by an anonymous individual. The data was collected at 5 minute intervals through out the day during the months of October and November 2012. Included is the number of steps taken in each 5 minute interval.

```{r init, results='hide'}
## create the directory if it does not exists. 
if (!file.exists("./data")) {
    dir.create("./data")
}

zipFileNm <- "./data/activity.zip"
dLoadFileNm <- "./data/sourceDt.RData"

## if the zip file does not exist download the file
if (!file.exists(zipFileNm)) {
    # this file stores the date when the archive was downloaded
    
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url, destfile = zipFileNm)
    
    # extract the archive
    unzip(zipFileNm, overwrite=TRUE, exdir="./data")
    
    # update the date the file was downloaded
    downloadDtTm <- format(Sys.time(), "%b %d %Y at %H:%M:%S")
    save(downloadDtTm, file = dLoadFileNm)        
} else {
    load(dLoadFileNm, envir=environment())    
}
```

## Loading and preprocessing the data

The data was downloaded from the url https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip. It was downloaded on **`r downloadDtTm`**. 

```{r loadProcData}
data <- read.csv("./data/activity.csv", header = TRUE)

# Convert the data field to a date
data$date <- as.Date(as.character(data$date), format="%Y-%m-%d")
```

The read data consists of `r nrow(data)` rows and `r ncol(data)` columns. The column names are **`r names(data)`** and their respective class is **`r sapply(data, class)`**.

```{r loadLibraryies}
# Install and load libraries used here
for (x in c("ggplot2", "plyr", "grid", "gridExtra")) {
    if (x %in% rownames(installed.packages()) == FALSE) {
        install.packages(x)
        }
    }
require(ggplot2)
require(grid)
require(gridExtra)
require(plyr)
```

## What is mean total number of steps taken per day?
```{r}
data1 <- data[complete.cases(data$steps),1:2]
plot1Data <- ddply(data1, .(date), summarize, totSteps = sum(steps))
p <- ggplot(plot1Data, aes(x=date, y = totSteps))
p <- p + geom_histogram(stat="identity")
p <- p + ylab("Total Steps") + xlab("Date")
p <- p + theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```

The mean of the total number of steps per day is **```r format(mean(plot1Data$totSteps), digits=2) ```**.
The median of the total number of steps per day is **```r format(median(plot1Data$totSteps), digits=2)```**.

##What is the average daily activity pattern?
```{r Chart2}
data2 <- data[complete.cases(data$steps),c(1,3)]
plot2Data <- ddply(data2, .(interval), summarize, avgSteps = mean(steps))

# Format the interval so that it is present as a time
plot2Data$interval <- sprintf("%04d", plot2Data$interval)
plot2Data$interval <- strptime(paste0(substr(plot2Data$interval, 1, 2), ":", substr(plot2Data$interval, 3, 4)), "%H:%M")
plot(plot2Data$interval, plot2Data$avgSteps, type="l", 
     xlab="Day interval", 
     ylab="Average Daily Steps", 
     main="Average Steps across the day")
```

The 5-minute interval, on average across all the days with the largest number of steps occurs at ***```r strftime(plot2Data$interval[which.max(plot2Data$avgSteps)], format="%H:%M")```***.

##Inputing missing values

```{r MissingValues}
# The number of rows with NA values can be computed by subtracting all the read rows from the number of complete cases computed for earlier parts of this solution.
missingVals <- nrow(data) - nrow(data1)
```

The number of missing values in the data is **```r missingVals```**.

```{r fillNAs}
# copy the dataset to a new one
filledData <- data

# Get the intervals that contain NA
intervalNA <- unique(filledData$interval[is.na(filledData$steps)])
for (x in intervalNA) {
    calcMean <- mean(filledData$steps[filledData$interval == x], na.rm=TRUE)
    filledData$steps[filledData$interval == x & is.na(filledData$steps)] <- calcMean
}
```

```{r Chart3a, fig.height=8}
plot3Data <- ddply(filledData, .(date), summarize, 
                   totSteps = sum(steps), 
                   meanSteps = mean(steps), 
                   medianSteps = median(steps))
p1 <- ggplot(plot3Data, aes(x=date, y = totSteps))
p1 <- p1 + geom_histogram(stat="identity")
p1 <- p1 + ylab("Total Steps") + xlab("Date") + ggtitle("Total steps")
p1 <- p1 + theme(axis.text.x = element_text(angle = 90, hjust = 1))

p2 <- ggplot(plot3Data, aes(x=date, y = meanSteps))
p2 <- p2 + geom_histogram(stat="identity")
p2 <- p2 + ylab("Mean Steps") + xlab("Date") + ggtitle("Mean steps")
p2 <- p2 + theme(axis.text.x = element_text(angle = 90, hjust = 1))

p3 <- ggplot(plot3Data, aes(x=date, y = medianSteps))
p3 <- p3 + geom_histogram(stat="identity")
p3 <- p3 + ylab("Median Steps") + xlab("Date") + ggtitle("Median steps")
p3 <- p3 + theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(p1, p2, p3, nrow = 3, main = "Analysis of step movement by day")
```

The charts above where generated from the data below:

```{r }
plot3Data
```

##Are there differences in activity patterns between weekdays and weekends?

```{r BuildWeekDayEnd}
# Indicate whether date is a week day or week end
filledData$dayType <- ifelse(weekdays(filledData$date) %in% c("Saturday", "Sunday"), 0, 1)

# and convert it into a factor
filledData$dayType <- factor(filledData$dayType, labels=c('Weekend', "Weekday"))

plot4Data <- ddply(filledData, .(interval, dayType), summarize, avgSteps = mean(steps))
ggplot(plot4Data, aes(x=interval, y = avgSteps)) +
    facet_grid(facets=(dayType ~ .)) +
    geom_line() + 
    xlab("5 Minute Interval") +
    ylab("Average Steps")
```

